---
title: "Differential expression and Venn diagrams (Adam's code) version 2"
output: html_document
---


Update: 7th November 2018

```{r, include=FALSE}


source('functions.R')
#library(org.Mm.eg.db)
#library(AnnotationDbi)
library(DESeq2)
library(pheatmap)
library(tidyverse)
library(DT)
library(limma)
library(pheatmap)
library(biomaRt)
library(dplyr)
#Have to use Biomart for gene annotation as more complete than Mm.eg.db (see Results chunk, unchanged in Volcano plots)
#Have to load Biomart before dplyr as overrides some functions!


```


```{r , echo=TRUE,warning=FALSE,message=FALSE,error=FALSE, include=FALSE}

df_mRNA <- read.table("genes.csv", sep = ",", header = TRUE, row.names=1)

meta_data <- read.table("meta_data.csv", sep=",", header = TRUE)

rownames(meta_data) <- meta_data$Sample

df_mRNA = df_mRNA[,rownames(meta_data)]

all(rownames(meta_data) %in% colnames(df_mRNA))


```


```{r , include = FALSE}
# set up meta_data for all experiments

meta_data <- meta_data %>% 
  filter(Cell_line == "A20") %>% 
  column_to_rownames("Sample")

```




```{r , message=FALSE, include=FALSE}
## Set up the DESeq2 object
res_e6.1 <- run_deseq2(as.data.frame(df_mRNA), meta_data, control = "neg", test="e6_1", value="Condition")

res_e6.2 <- run_deseq2(as.data.frame(df_mRNA), meta_data, control = "neg", test="e6_2", value="Condition")

res_e7.1 <- run_deseq2(as.data.frame(df_mRNA), meta_data, control = "neg", test="e7_1", value="Condition")

res_e7.2 <- run_deseq2(as.data.frame(df_mRNA), meta_data, control = "neg", test="e7_2", value="Condition")


```

The following describes the analysis of the data using DEseq2. I have separated the analyses into the four different contrasts and have plotted summary statistics and MA plots for each.  


## Summary of the data {.tabset .tabset-fade}

### A20 Negative vs A20 Foxp1 Clone E6.1

```{r , echo=FALSE}
summary(res_e6.1, alpha=0.05)
```

### A20 Negative vs A20 Foxp1 Clone E6.2

```{r , echo=FALSE}
summary(res_e6.2, alpha=0.05)
```

### A20 Negative vs A20 Foxp1 Clone E7.1

```{r , echo=FALSE}
summary(res_e7.1, alpha=0.05)
```


### A20 Negative vs A20 Foxp1 Clone E7.2

```{r , echo=FALSE}
summary(res_e7.2, alpha=0.05)
```

## MA plots {.tabset .tabset-fade}

In DESeq2, the function plotMA shows the log2 fold changes attributable to a given variable over the mean of normalized counts for all the samples in the DESeqDataSet. Points will be colored red if the adjusted p value is less than 0.01. Points which fall out of the window are plotted as open triangles pointing either up or down.

### A20 Negative vs A20 Foxp1 Clone E6.1

```{r, echo=FALSE}
DESeq2::plotMA(res_e6.1)
```

### A20 Negative vs A20 Foxp1 Clone E6.2

```{r, echo=FALSE}
DESeq2::plotMA(res_e6.2)
```

### A20 Negative vs A20 Foxp1 Clone E7.1

```{r, echo=FALSE}
DESeq2::plotMA(res_e7.1)
```

### A20 Negative vs A20 Foxp1 Clone E7.2

```{r, echo=FALSE}
DESeq2::plotMA(res_e7.2)
```



## Results tables

The folowing results tables show the significant genes. Filtering has been performed with a log2 fold change +/- 2.


### A20 Negative vs A20 Foxp1 Clone E6.1

```{r , message=FALSE, echo=FALSE}


dt <- filter_biomart_genes(as.data.frame(res_e6.1), name="E6_1")


datatable(dt$sig)
```


### A20 Negative vs A20 Foxp1 Clone E6.2

```{r , message=FALSE, echo=FALSE}
dt <- filter_biomart_genes(as.data.frame(res_e6.2), name="E6_2")


datatable(dt$sig)
```


### A20 Negative vs A20 Foxp1 Clone E7.1

```{r , message=FALSE, echo=FALSE}
dt <- filter_biomart_genes(as.data.frame(res_e7.1), name="E7_1")


datatable(dt$sig)
```



### A20 Negative vs A20 Foxp1 Clone E7.2

```{r , message=FALSE, echo=FALSE}
dt <- filter_biomart_genes(as.data.frame(res_e7.2), name="E7_2")


datatable(dt$sig)
```




```{r, echo=FALSE}

# Merge the data for reactome
E6_1 <- as.data.frame(res_e6.1)
E6_2 <- as.data.frame(res_e6.2)
E7_1 <- as.data.frame(res_e7.1)
E7_2 <- as.data.frame(res_e7.2)

# downregulated

E6_1_down <- E6_1 %>% 
  rownames_to_column("ensembl") %>% 
  filter(log2FoldChange < -1)

E6_2_down <- E6_2 %>% 
  rownames_to_column("ensembl") %>% 
  filter(log2FoldChange < -1)

E7_1_down <- E7_1 %>% 
  rownames_to_column("ensembl") %>% 
  filter(log2FoldChange < -1)

E7_2_down <- E7_2 %>% 
  rownames_to_column("ensembl") %>% 
  filter(log2FoldChange < -1)

# upregulated

E6_1_up <- E6_1 %>% 
  rownames_to_column("ensembl") %>%
  filter(log2FoldChange > 1)

E6_2_up <- E6_2 %>% 
  rownames_to_column("ensembl") %>%
  filter(log2FoldChange > 1)

E7_1_up <- E7_1 %>% 
  rownames_to_column("ensembl") %>%
  filter(log2FoldChange > 1)

E7_2_up <- E7_2 %>% 
  rownames_to_column("ensembl") %>%
  filter(log2FoldChange > 1)

```


# Shared downregulated genes in A20 Foxp1 E6 clones

```{r, message=FALSE, echo=FALSE}
library(reshape)
library(reshape2)
library(wesanderson)
library(dplyr)


E6_1_down <- E6_1_down %>% 
  select(ensembl, log2FoldChange)
E6_2_down <- E6_2_down %>% 
  select(ensembl, log2FoldChange)

merge_down <- merge(E6_1_down, E6_2_down, by="ensembl", all = TRUE)

merge_down <- merge_down %>% 
  mutate(E6_1_LFC = log2FoldChange.x, E6_2_LFC = log2FoldChange.y) %>% 
  select(-log2FoldChange.x, -log2FoldChange.y)

merge_down[is.na(merge_down)] <- 0

e6.1 <- (merge_down$E6_1_LFC < -1)
e6.2 <- (merge_down$E6_2_LFC < -1)

c2 <- cbind(e6.1, e6.2)

a <- vennCounts(c2)
a

library("RColorBrewer")
library(wesanderson)
c("red", "blue", "darkgreen", "yellow")

tiff("venn2/e6_downregulated_venn.tiff")
vennDiagram(a, circle.col = wes_palette("Darjeeling1"))
#dev.off()
knitr::include_graphics("venn2/e6_downregulated_venn.tiff")



# Next subset the overlapping genes for all

merge_down <- ensembl_to_symbol_biomart(merge_down, "ensembl")
down <- merge_down[complete.cases(merge_down),]


write.csv(down, file="venn2/e6_down.csv")
```


# Shared downregulated genes in A20 Foxp1 E7 clones

```{r, echo=FALSE}
library(reshape)
library(reshape2)
library(dplyr)

E7_1_down <- E7_1_down %>% 
  select(ensembl, log2FoldChange)
E7_2_down <- E7_2_down %>% 
  select(ensembl, log2FoldChange)

merge_down <- merge(E7_1_down, E7_2_down, by="ensembl", all = TRUE)

merge_down <- merge_down %>% 
  mutate(E7_1_LFC = log2FoldChange.x, E7_2_LFC = log2FoldChange.y) %>% 
  select(-log2FoldChange.x, -log2FoldChange.y)

merge_down[is.na(merge_down)] <- 0

e7.1 <- (merge_down$E7_1_LFC < -1)
e7.2 <- (merge_down$E7_2_LFC < -1)

c2 <- cbind(e7.1, e7.2)

a <- vennCounts(c2)
a

library("RColorBrewer")
library(wesanderson)
c("red", "blue", "darkgreen", "yellow")

#setEPS()
#postscript("sum_downregulated_venn.eps")
#vennDiagram(a, circle.col = wes_palette("Darjeeling1"))
#dev.off()

tiff("venn2/e7_downregulated_venn.tiff")
vennDiagram(a, circle.col = wes_palette("Darjeeling1"))
#dev.off()
knitr::include_graphics("venn2/e7_downregulated_venn.tiff")

# Next subset the overlapping genes for all

merge_down <- ensembl_to_symbol_biomart(merge_down, "ensembl")
down <- merge_down[complete.cases(merge_down),]


write.csv(down, file="venn2/e7_down.csv")
```

# Shared downregulated genes in all Foxp1 clones

```{r, echo=FALSE}
library(reshape)
library(reshape2)
library(dplyr)

E6_1_down <- E6_1_down %>% 
  select(ensembl, log2FoldChange)
E6_2_down <- E6_2_down %>% 
  select(ensembl, log2FoldChange)

E7_1_down <- E7_1_down %>% 
  select(ensembl, log2FoldChange)

E7_2_down <- E7_2_down %>% 
  select(ensembl, log2FoldChange)

merge_down <- merge(E6_1_down, E6_2_down, by="ensembl", all = TRUE)

merge_down <- merge_down %>% 
  mutate(E6_1_LFC = log2FoldChange.x, E6_2_LFC = log2FoldChange.y) %>% 
  select(-log2FoldChange.x, -log2FoldChange.y)

merge_down <- merge(merge_down, E7_1_down, by="ensembl", all = TRUE)
merge_down <- merge_down %>% 
  mutate(E7_1_LFC = log2FoldChange) %>% 
  select(-log2FoldChange)

merge_down <- merge(merge_down, E7_2_down, by="ensembl", all = TRUE)
merge_down <- merge_down %>% 
  mutate(E7_2_LFC = log2FoldChange) %>% 
  select(-log2FoldChange)


merge_down[is.na(merge_down)] <- 0


e6.1 <- (merge_down$E6_1_LFC < -1)
e6.2 <- (merge_down$E6_2_LFC < -1)
e7.1 <- (merge_down$E7_1_LFC < -1)
e7.2 <- (merge_down$E7_2_LFC < -1)

c2 <- cbind(e6.1, e6.2, e7.1, e7.2)

a <- vennCounts(c2)
a

library("RColorBrewer")
library(wesanderson)
c("red", "blue", "darkgreen", "yellow")


tiff("venn2/all_downregulated_venn.tiff")
vennDiagram(a, circle.col = wes_palette("Darjeeling1"))
#dev.off()
knitr::include_graphics("venn2/all_downregulated_venn.tiff")

# Next subset the overlapping genes for all

merge_down <- ensembl_to_symbol_biomart(merge_down, "ensembl")
down <- merge_down[complete.cases(merge_down),]


write.csv(down, file="venn2/all_down.csv")
```


# Shared upregulated genes in A20 Foxp1 E6 clones

```{r, echo=FALSE}

E6_1_up <- E6_1_up %>% 
  select(ensembl, log2FoldChange)
E6_2_up <- E6_2_up %>% 
  select(ensembl, log2FoldChange)

merge_up <- merge(E6_1_up, E6_2_up, by="ensembl", all = TRUE)

merge_up <- merge_up %>% 
  mutate(E6_1_LFC = log2FoldChange.x, E6_2_LFC = log2FoldChange.y) %>% 
  select(-log2FoldChange.x, -log2FoldChange.y)

merge_up[is.na(merge_up)] <- 0

e6.1 <- (merge_up$E6_1_LFC > 1)
e6.2 <- (merge_up$E6_2_LFC > 1)

c2 <- cbind(e6.1, e6.2)

a <- vennCounts(c2)
a

library("RColorBrewer")
library(wesanderson)
c("red", "blue", "darkgreen", "yellow")

tiff("venn2/e6_upregulated_venn.tiff")
vennDiagram(a, circle.col = wes_palette("Darjeeling1"))
#dev.off()
knitr::include_graphics("venn2/e6_upregulated_venn.tiff")

# Next subset the overlapping genes for all

merge_up <- ensembl_to_symbol_biomart(merge_up, "ensembl")
up <- merge_up[complete.cases(merge_up),]


write.csv(up, file="venn2/e6_up.csv")


```

# Shared upregulated genes in A20 Foxp1 E7 clones

```{r, echo=FALSE}

E7_1_up <- E7_1_up %>% 
  select(ensembl, log2FoldChange)
E7_2_up <- E7_2_up %>% 
  select(ensembl, log2FoldChange)

merge_up <- merge(E7_1_up, E7_2_up, by="ensembl", all = TRUE)

merge_up <- merge_up %>% 
  mutate(E7_1_LFC = log2FoldChange.x, E7_2_LFC = log2FoldChange.y) %>% 
  select(-log2FoldChange.x, -log2FoldChange.y)

merge_up[is.na(merge_up)] <- 0

e7.1 <- (merge_up$E7_1_LFC > 1)
e7.2 <- (merge_up$E7_2_LFC > 1)

c2 <- cbind(e7.1, e7.2)

a <- vennCounts(c2)
a

library("RColorBrewer")
library(wesanderson)
c("red", "blue", "darkgreen", "yellow")

tiff("venn2/e7_upregulated_venn.tiff")
vennDiagram(a, circle.col = wes_palette("Darjeeling1"))
#dev.off()
knitr::include_graphics("venn2/e7_upregulated_venn.tiff")

# Next subset the overlapping genes for all

merge_up <- ensembl_to_symbol_biomart(merge_up, "ensembl")
up <- merge_up[complete.cases(merge_up),]


write.csv(up, file="venn2/e7_up.csv")
```

# Shared upregulated genes in all Foxp1 clones

```{r, echo=FALSE}

E6_1_up <- E6_1_up %>% 
  select(ensembl, log2FoldChange)
E6_2_up <- E6_2_up %>% 
  select(ensembl, log2FoldChange)
E7_1_up <- E7_1_up %>% 
  select(ensembl, log2FoldChange)
E7_2_up <- E7_2_up %>% 
  select(ensembl, log2FoldChange)

merge_up <- merge(E6_1_up, E6_2_up, by="ensembl", all = TRUE)

merge_up <- merge_up %>% 
  mutate(E6_1_LFC = log2FoldChange.x, E6_2_LFC = log2FoldChange.y) %>% 
  select(-log2FoldChange.x, -log2FoldChange.y)

merge_up <- merge(merge_up, E7_1_up, by="ensembl", all = TRUE)
merge_up <- merge_up %>% 
  mutate(E7_1_LFC = log2FoldChange) %>% 
  select(-log2FoldChange)

merge_up <- merge(merge_up, E7_2_up, by="ensembl", all = TRUE)
merge_up <- merge_up %>% 
  mutate(E7_2_LFC = log2FoldChange) %>% 
  select(-log2FoldChange)

merge_up[is.na(merge_up)] <- 0

e6.1 <- (merge_up$E6_1_LFC > 1)
e6.2 <- (merge_up$E6_2_LFC > 1)
e7.1 <- (merge_up$E7_1_LFC > 1)
e7.2 <- (merge_up$E7_2_LFC > 1)

c2 <- cbind(e6.1, e6.2, e7.1, e7.2)

a <- vennCounts(c2)
a

library("RColorBrewer")
library(wesanderson)
c("red", "blue", "darkgreen", "yellow")

tiff("venn2/all_upregulated_venn.tiff")
vennDiagram(a, circle.col = wes_palette("Darjeeling1"))
#dev.off()
knitr::include_graphics("venn2/all_upregulated_venn.tiff")

# Next subset the overlapping genes for all

merge_up <- ensembl_to_symbol_biomart(merge_up, "ensembl")
up <- merge_up[complete.cases(merge_up),]


write.csv(up, file="venn2/all_up.csv")
```

